\tit

### 2021-02-28 paper.tex
fix wonky footnote; trim some words

Since {\AB{u} \AF ⊙ \AIC{[]} \AD ≡ \AB{u}}, it follows that\footnote{For functions \AB{f} and \AB{g}, {\AB{f} \AF{≗} \AB{g}} is extensional equality, i.e., {\AS ∀ \AB x \AS → \AB{f} \AB{x} \AD{≡} \AB{g} \AB{x}}.}
\agda{ν∘𝒟}
Moreover, the function \AF{𝒟} distributes over \AIC{[]} and \AF{\_⊙\_}:\footnote{In algebraic terms, argument-flipped \AF{𝒟} is a monoid homomorphism from lists to endofunctions.}
\agda{𝒟[]⊙}
These facts suggest that \AB{𝒟} can be computed one list element at a time via a more specialized operation, as is indeed the case:\footnote{Repeated \AF{δ} application is expressed here via a standard left fold\out{ \stdlibCitep{Data.List}}:
\agda{foldl}\vspace{-2ex}
In \AF{𝒟foldl}, {\AB{X} \AS = \AB{A} \AF ✶ \AS → \AB B}.
}
\agda{δ}
%\vspace{-4ex}
\agda{𝒟foldl}
Each use of \AF{δ} thus takes us one step closer to reducing general language membership to nullability.
%% In other words,%
%% \agda{ν∘foldlδ}
This simple observation is the semantic heart of the syntactic technique of \emph{derivatives of regular expressions} as used for efficient recognition of regular languages \citep{Brzozowski64}, later extended to parsing general context-free languages \citep{Might2010YaccID}.
Lemmas \AF{ν∘𝒟} and \AF{𝒟foldl} liberate this technique from the assumption that languages are represented \emph{symbolically}, by some form of grammar (e.g., regular or context-free), inviting other representations as we will see in \secref{Automatic Differentiation}.

@@

For an effective implementation, we must bridge the gap between a type (of membership proofs) and its decidable inhabitation.
Fortunately, there is a convenient and compositional way to do so.
For any type \AB{A}, the type {\AF{Dec} \AB{A}} contains proof of \AB{A} or a proof of {\AF ¬ \AB{A}} (defined as usual to mean {\AB{A} \AS → \AD{⊥}}):\footnote{This simple \AF{Dec} definition been superceded by a more efficient but more complex version \stdlibCitep{Relation.Nullary}.}
\agda{Dec}
For compositionality, we will use a few operations that lift decidability of types to decidability of constructions on those types, as shown in \figrefdef{compositional-dec}{Compositional decidability}{\agda{compositional-dec}}.\footnote{%
A few explanatory remarks:
\begin{itemize}

>>>>

Since {\AB{u} \AF ⊙ \AIC{[]} \AD ≡ \AB{u}}, it follows that\footnote{For functions \AB{f} and \AB{g}, {\AB{f} \AF{≗} \AB{g}} is extensional equality, i.e., {\AS ∀ \AB x \AS → \AB{f} \AB{x} \AD{≡} \AB{g} \AB{x}}.}
\agda{ν∘𝒟}
Moreover, the function \AF{𝒟} distributes over \AIC{[]} and \AF{\_⊙\_}:\footnote{In algebraic terms, argument-flipped \AF{𝒟} is a monoid homomorphism from lists to endofunctions.}
\agda{𝒟[]⊙}
These facts suggest that \AB{𝒟} can be computed one list element at a time via a more specialized operation, as is indeed the case:\footnote{Repeated \AF{δ} application is expressed here via a standard left fold (with {\AB{X} \AS = \AB{A} \AF ✶ \AS → \AB B} in \AF{𝒟foldl}):
\agda{foldl}
}
\agda{δ}
%\vspace{-4ex}
\agda{𝒟foldl}
Each use of \AF{δ} thus takes us one step closer to reducing general language membership to nullability.
%% In other words,%
%% \agda{ν∘foldlδ}
This simple observation is the semantic heart of the syntactic technique of \emph{derivatives of regular expressions} as used for efficient recognition of regular languages \citep{Brzozowski64}, later extended to parsing general context-free languages \citep{Might2010YaccID}.
Lemmas \AF{ν∘𝒟} and \AF{𝒟foldl} liberate this technique from the assumption that languages are represented \emph{symbolically}, by some form of grammar (e.g., regular or context-free), inviting other representations as we will see in \secref{Automatic Differentiation}.

@@

For effective implementations, we must bridge the gap between a type (of membership proofs) and its decidable inhabitation.
Fortunately, there is a convenient and compositional way to do so.
For type \AB{A}, the type {\AF{Dec} \AB{A}} contains proof of \AB{A} or a proof of {\AF ¬ \AB{A}} (defined as usual to mean {\AB{A} \AS → \AD{⊥}}):\footnote{This simple \AF{Dec} definition been superceded by a more efficient but more complex version \stdlibCitep{Relation.Nullary}.}
\agda{Dec}
For compositionality, we will use a few operations that lift decidability of types to decidability of constructions on those types, as shown in \figrefdef{compositional-dec}{Compositional decidability}{\agda{compositional-dec}}.\footnote{%
A few explanatory remarks:
\begin{itemize}
